Plant trait cleaning and combining

### 2024-06-11 code/L1/L1_P4_TropicalAndes_plant_traits.Rmd
add output, summary graphs, save figures into figure folder

---
title: "Plant trait cleaning and combining"
author: "Hazel J. Anderson"
project: "Plant-Frugivore Diversity"
collaborators: "Beth E. Gerstner, Phoebe L. Zarnetske"
overview: "This script subsets plant trait data by synonyms species list and combine."
data input: "TropicalAndes_all_plant_traits_harmonized_subset.csv"
data output: "ADD"
date: "2023-11-04"
output: html_document
---

# Load required packages
```{r}
library(dplyr)
library(tidyr)
library(purrr)
library(rstatix)
library(funbiogeo)
library(visdat)
```

@@

@@

## Plant lifespan/longevity
```{r}
traits_plantLifespan <- traits %>%
  filter(grepl('Plant lifespan \\(longevity\\)|longest whole plant longevity|maximum whole plant longevity|plant_lifespan', TraitName))
traits_plantLifespan$OrigTraitValue <- traits_plantLifespan$TraitValue

@@

## Seed mass
```{r}
traits_seedMass <- traits %>%
  filter(grepl('Seed dry mass|seed mass|seed_mass_max|seed_mass_mean|seed_mass_min', TraitName))
traits_seedMass$OrigTraitValue <- traits_seedMass$TraitValue

@@

## Seed length
```{r}
traits_seedLength <- traits %>%
  filter(grepl('Seed length|seed_length_max|seed_length_mean|seed_length_min', TraitName))
traits_seedLength$OrigTraitValue <- traits_seedLength$TraitValue

@@

## Seed width
```{r}
traits_seedWidth <- traits %>%
  filter(grepl('Seed width|seed_width_max|seed_width_mean|seed_width_min'
, TraitName))
traits_seedWidth$OrigTraitValue <- traits_seedWidth$TraitValue

@@

# Combine all traits
```{r}
trait_dfs <- list(averagetraits_fruitLength, averagetraits_fruitMass, averagetraits_plantHeight, averagetraits_plantLifespan, averagetraits_seedLength, averagetraits_seedMass, averagetraits_seedWidth, averagetraits_dispersal, averagetraits_growthForm, averagetraits_fruitType)

# Function to combine trait dataframes only matching columns
combine_matching_columns <- function(df_list) {
  # Find the common columns across all data frames
  common_cols <- Reduce(intersect, lapply(df_list, colnames))

@@

# Transform plant dataframe from long to wide
```{r}
wide_traits <- spread(clean_traits, key = TraitName, value = TraitValue)

#remove X column from dataframe
wide_plant_traits <- wide_traits[, !colnames(wide_traits) %in% c("X", "records_used", "variance"), drop = FALSE]
```

@@

```{r}
fb_plot_species_traits_completeness(wide_plant_traits)
fb_plot_number_species_by_trait(wide_plant_traits)
fb_table_trait_summary(wide_plant_traits)
```

```{r}
vis_dat(wide_plant_traits)
vis_miss(wide_plant_traits)
```

# Write data to csv
```{r}
write.csv(wide_plant_traits, file.path(output_path,"TropicalAndes_all_plant_traits_standardized.csv"))
write.csv(clean_traits, file.path(output_path,"TropicalAndes_all_plant_traits_standardized_with_recordCount.csv"))
```

>>>>

---
title: "Plant trait cleaning and combining"
author: "Hazel J. Anderson"
project: "Plant-Frugivore Diversity"
collaborators: "Beth E. Gerstner, Phoebe L. Zarnetske"
overview: "This script subsets plant trait data by synonyms species list and combine."
data input: "TropicalAndes_all_plant_traits_harmonized_subset.csv"
data output: "TropicalAndes_all_plant_traits_standardized.csv, TropicalAndes_all_plant_traits_standardized_with_recordCount.csv"
date: "2023-11-04"
output: 
  html_document:
    code-link: true
    df_print: paged
---

# Load required packages
```{r}
library(dplyr)
library(tidyr)
library(purrr)
library(rstatix)
library(funbiogeo)
library(visdat)
library(ggplot2)
```

@@

```{r}
# remove X column
plant_traits <- plant_traits[, !colnames(plant_traits) %in% "X", drop = FALSE]

# Convert all character columns to UTF-8 encoding to handle encoding issues
plant_traits <- plant_traits %>%
    mutate(across(where(is.character), ~ iconv(., from = "", to = "UTF-8", sub = "byte")))

# Replace the specific value in TraitName column
plant_traits <- plant_traits %>%
  mutate(TraitName = gsub("Fruit texture inside \\(fleshy, leathrey, .+\\)", "Fruit texture inside", TraitName))
```

@@

## Plant lifespan/longevity
```{r}
traits_plantLifespan <- plant_traits %>%
  filter(grepl('Plant lifespan \\(longevity\\)|longest whole plant longevity|maximum whole plant longevity|plant_lifespan', TraitName))
traits_plantLifespan$OrigTraitValue <- traits_plantLifespan$TraitValue

@@

## Seed mass
```{r}
traits_seedMass <- plant_traits %>%
  filter(grepl('Seed dry mass|seed mass|seed_mass_max|seed_mass_mean|seed_mass_min', TraitName))
traits_seedMass$OrigTraitValue <- traits_seedMass$TraitValue

@@

## Seed length
```{r}
traits_seedLength <- plant_traits %>%
  filter(grepl('Seed length|seed_length_max|seed_length_mean|seed_length_min', TraitName))
traits_seedLength$OrigTraitValue <- traits_seedLength$TraitValue

@@

## Seed width
```{r}
traits_seedWidth <- plant_traits %>%
  filter(grepl('Seed width|seed_width_max|seed_width_mean|seed_width_min'
, TraitName))
traits_seedWidth$OrigTraitValue <- traits_seedWidth$TraitValue

@@

# Combine all traits
```{r}
# for summary
unaveraged_traits <- rbind(traits_dispersal, traits_fruitLength, traits_fruitMass, traits_fruitType, traits_growthForm, traits_plantHeight, traits_plantLifespan, traits_seedLength, traits_seedMass, traits_seedWidth)
```

```{r}
trait_dfs <- list(averagetraits_fruitLength, averagetraits_fruitMass, averagetraits_plantHeight, averagetraits_plantLifespan, averagetraits_seedLength, averagetraits_seedMass, averagetraits_seedWidth, averagetraits_dispersal, averagetraits_growthForm, averagetraits_fruitType)

# Function to combine trait dataframes only matching columns
combine_matching_columns <- function(df_list) {
  # Only use specified columns
  common_cols <- Reduce(intersect, lapply(df_list, colnames))

@@

# Transform plant dataframe from long to wide
```{r}
wide_plant_traits <- clean_traits %>%
  select(Accepted_species, TraitName, TraitValue) %>%
  pivot_wider(names_from = TraitName, values_from = TraitValue)
nrow(wide_plant_traits)
```

@@

```{r}
fb_plot_species_traits_completeness(wide_plant_traits)
ggsave("plant_trait_completeness.png", plot = last_plot(), path = figure_path)
fb_plot_number_species_by_trait(wide_plant_traits)
ggsave("plant_number_species_trait.png", plot = last_plot(), path = figure_path)
fb_table_trait_summary(wide_plant_traits)
```

```{r}
vis_dat(wide_plant_traits)
ggsave("plant_trait_visual_type.png", plot = last_plot(), path = figure_path)
vis_miss(wide_plant_traits)
ggsave("plant_trait_visual_missing.png", plot = last_plot(), path = figure_path)
```

```{r}
# Count the number of traits per DatabaseSource and TraitName
trait_counts <- unaveraged_traits %>%
  count(DatabaseSource, TraitName)

# Create the faceted bar plot with free y-axis scales
ggplot(trait_counts, aes(x = DatabaseSource, y = n, fill = DatabaseSource)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Trait Counts per Database",
       x = "Trait Name",
       y = "Trait Count",
       fill = "Database Source") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_wrap(~ TraitName, scales = "free_y")
ggsave("plant_trait_counts_per_database.png", plot = last_plot(), path = figure_path)
```

# Write data to csv
```{r}
write.csv(wide_plant_traits, file.path(output_path,"TropicalAndes_all_plant_traits_standardized.csv"))
write.csv(clean_traits, file.path(output_path,"TropicalAndes_all_plant_traits_standardized_with_recordCount.csv"))
write.csv(unaveraged_traits, file.path(output_path,"TropicalAndes_all_plant_traits_cleaned_unaveraged.csv"))
```

# Package citations and session info
```{r}
library(report)
report::cite_packages()

devtools::session_info()
```
